### 2022-03-29 pre/pre.tex
biblio
more

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We finish with the
    description of the superposition calculus. As this thesis discusses practical
    aspects of theorem proving, we define only the fundamental notions, while
    more advanced notions are intuitively described with references to rigorous
    definitions. The text of this chapter is
    partially based on the preliminaries sections of the publications listed in
    Chapter \ref{ch:intro}.

\end{abstract}

@@

Terms are used to build \emph{atoms}. An atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable, then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for any value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic, it does not suffice to assign
values to variables to determine the truth value of the formula. It is also
necessary to interpret the symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

Even though we do not formally define the semantics of logic, we assume the natural
extensions of domain, valuation, interpretation and model (as defined by Fitting
\cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
of (un)satisfiability and (in)validity are assumed. We write
$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
(i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
$\ourmodel \models N.$  Abusing notation, we write $M \models N$ to denote that
$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$

A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
We say a term is {\em ground} if it has no variables. By $u[s]$
we denote that $s$ is a subterm of $u$.

@@

In this thesis we use classic extensional simply typed monomorphic higher-order
logic with the choice operator, but without the description operator, the axiom of infinity,
and the axiom of at least two individuals \cite[Sect.~3.5]{bm-14-automation-ho}.
This logic corresponds to Henkin's extensional type theory. We additionally use
some variations of this logic, clearly stating the differences. Assuming a set of
base types $T$ and a set of symbols $\Sigma$, let us define terms and types.
Types are either base types $\tau \in T$, or function types $\tau \rightarrow
\upsilon$ where both $\tau$ and $\upsilon$ are types. When we write $\tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \upsilon$ we assume $\upsilon$
to be a base type ($\upsilon \in T$), by convention. Each symbol $\cst{f} \in
\Sigma$ is assigned a type.

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau \rightarrow
\upsilon$ and $\tau$, respectively, then $s \, t$ is a term of type $\upsilon$.
Lastly, if $x$ is a bound variable of type $\tau$ and $s$ is a term of type
$\upsilon$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau
\rightarrow \upsilon$. The syntactic distinction between free and bound
variables gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
\cst{a}}$) \cite{tn-93-patterns}. Note that there is no requirement that a
symbol is not applied to terms of Boolean type as in first-order logic.
Furthermore, all logical connectives are symbols present in the set $\Sigma$.
This means that higher-order logic does not distinguish between terms and
formulas. For example, $\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand
\cst{q})$ are well-formed terms of higher-order logic. For convenience, we still
call terms of Boolean type \emph{formulas}. Our logic also supports Hilbert
choice operator $\varepsilon$. Intuitively, this operator chooses an arbitrary value that
satisfies a predicate and it is axiomatized as follows: $\iforall p. \, (\iexists
x. \, p \, x) \iimplies p \, (\varepsilon \, p)$.

\looseness=-1
With $\alpha$-conversion we assume the rule $(\lambda x. \, s) \longrightarrow
\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur loosely bound in $s$
and $y$ is not bound in $s$. Application of a substitution to a term,
implicitly $\alpha$-renames bound variables to avoid capture.  For example, $\{X
\mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
$\beta$-reduction rule, roughly speaking, corresponds to passing arguments in a
function call: $(\lambda x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where
the bound variables in $s$ are renamed to avoid capture. Lastly,
$\eta$-reduction is defined as $\lambda x.\, s \, x \longrightarrow s$ under the
condition that $x$ is not loosely bound in $x$. We write $s \equi t$ if terms
$s$ and $t$ are equal modulo $\alpha\beta\eta$-conversion. Terms that are
convertible with respect to any of the rules form equivalence classes. Thus, we
also say that $\alpha$($\beta\eta$)-convertible terms are
$\alpha$($\beta\eta$)-equivalent.

@@

We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$,
omitting the length $n \geq 0$ when it is not important or it can be inferred. Every
$\beta$-reduced term can be written as $\lam{\tuple{x}{m}}{a \, \tuplen{t}}$,
where $a$ is not an application; we call $a$ the \emph{head} of the term. By
convention, $a, b$ and $\zeta$ denote heads. If $a$ is a free variable, we call the term
\emph{flex}; otherwise, the term is \emph{rigid}.

Deviating from the standard notion of higher-order subterm, we define subterms
on $\beta$-reduced terms as follows: a term $t$ is a subterm of $t$ at position
$\varepsilon$. If $s$ is a subterm of $u_i$ at position $p$, then $s$ is a
subterm of $a\;\overline{u}_n$ at position $i.p$. If $s$ is a subterm of $t$ at
position $p$, then $s$ is a subterm of $\lambda x.\, t$ at position $1.p$. Our
definition of subterm gracefully generalizes the corresponding first-order
notion: $\cst{a}$ is a subterm of $\cst{f} \, \cst{a} \, \cst{b}$ at position 1,
but $\cst{f}$ and $\cst{f} \, \cst{a}$ are not subterms of $\cst{f} \, \cst{a}
\, \cst{b}$. We say a term is ground if it has no free variables, and closed
if it has no loosely bound variables.

Throughout this thesis, we consider completeness of higher-order calculi only
with respect to Henkin semantics \cite{bm-14-automation-ho}. Note that no
complete calculus exists for higher-order calculi with standard (full)
semantics.

@@

The resolution and superposition calculi do not work directly on formulas, but
on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed into a set of clauses. For all three
discussed logics, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}.

\looseness=-1
An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em
literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq
t$. A clause $C$ is a finite multiset of literals, interpreted and written
disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
implicitly universally quantified. Note that clause-level operators ($\llor, \eq, \noteq$) are not typeset in bold
to distinguish them from term-building symbols ($\ior, \ieq, \ineq, \iand, \iforall, \ldots$) which are typeset in bold. In standard,
non-clausal first-order logic, an atom is either predicate or equational. For
uniformity, and to stay close to the implementation, we encode predicate atoms
as equations with $\itrue$. Negative literals are encoded as disequations. For
example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$. To lighten the
notation, we sometimes write the predicate literals in non-encoded form.
Applying a substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. We say
$\sigmacl{C}$ is a ground instance of $C$ if $\sigma$ is a grounding
substitution.
\pagebreak[2]

@@

This calculus works on problems in clausified form and proves the problem by
refuting its negation. In practice this means that, to apply superposition to a
given problem, we first must negate its conjecture and clausify the whole
problem using some of the clausification algorithms
\cite{nw-01-small-cnf,rsv-16-vcnf}. The calculus then applies its inference
rules to the clauses, adding results of the inferences to the working set of
clauses. As this process is quite prolific it also uses the term order to determine
if some clauses can be simplified or eliminated from the working set.
Superposition is refutationally complete: This means that, if results of all
inferences are computed in a systematic, fair manner, empty clause ($\bot$) will
eventually be derived for a provable problem. As first-order logic is
undecidable, no guarantees are given for unprovable problems.
Before introducing the rules of the calculus, we provide definitions of these and other
background concepts.

\subsection{Term Order and Selection}
\label{sec:pre:order}
Superposition calculus is parameterized by a {\em reduction ordering} $\succ$, an
ordering on terms that has the following properties:

\noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
    \textit{Irreflexive} & For all terms $s$, $s \not\succ s$ \\
    \textit{Transitive} & For all terms $s, t, u$,  if $s\succ t$ and $t \succ u$ then $s \succ u$ \\ 
    \textit{Subterm property} & For all terms $s$ and contexts $C$, $C[s] \succ s$ \\
    \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
    \textit{Respects contexts} & For all terms $s,t$ and contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
    \textit{Ground total} & For all distinct
    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$. \\
    \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
\end{tabular}

The term order is lifted to literals and clauses using the {\em multiset extension} of
$\succ$. The order is extended to multisets as follows
\cite[Sect.~2.5]{bg-01-resolution}. For two multisets $S_1$ and $S_2$, we write $S_1 \succ S_2$ if $S_1 \not= S_2$
and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
> S_2(y)$. We use the notation $S(x)$ to denote the number of occurrences of $x$ in
$S$. To use the  multiset extension, positive literals are represented as $ \{ \{s\},
\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
multisets of such literals.

Two orders that are commonly used in superposition theorem proving are the
Knuth-Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and the
lexicographic path order (LPO) \cite[Sect.~5.4.2]{bn-98-tr-and-all-that}.
KBO assigns integer weights to symbols and uses precedence between symbols to
break eventual ties. LPO entirely relies on precedence and inspects term
structures more closely to determine the order.

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
We always specify if the constraint should be interpreted as ordered or unordered.
We say a substitution $\sigma$ is \emph{more general} than $\varrho$ if there exists a substitution $\theta$
such that $\varrho = \theta\sigma$.
A \emph{most general unifier} is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
the most general unifier is unique up to variable renaming. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

@@

A (higher-order) unifier of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma(s) \equi \sigma(t)$, for all $s \unif t
\in E$. A complete set of unifiers of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier
corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist algorithms
that enumerate all elements of the CSU for unifiable terms.

These higher-order concepts gracefully generalize the corresponding first-order
ones. For example, if $s$ at $t$ are first-order, then there must exist a singleton
CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
unifiers as first-order or higher-order and ensure the context provides enough
information.

@@

Before spelling out the inference rules of standard, first-order superposition
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote the subterm of $t$ at position $p$ and with $t[p \leftarrow
t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within the literals in
\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.  With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using a horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

@@

\medskip
Let us give some intuition behind how these rules are applied in practice. The
\infname{ER} rule is used to replace equality with equality within a clause.
Consider the formula $\iforall x.\,  x \eq \cst{2} \iimplies \cst{even}(x)$. In
clausal form it is $X \noteq \cst{2} \llor \cst{even}(X)$. Applying the
\infname{ER} rule effectively applies the precondition $X \eq \cst{2}$ to the
conclusion $\cst{even}(X)$ to obtain $\cst{even}(\cst{2})$. The \infname{EF}
rule can be used to find instances of the clause that can be further simplified.
Consider the clause $\cst{f}(X) \eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor
\cst{p}(X, Y)$. Applying \infname{EF} to its first two literals yields $\cst{a}
\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ (with $\sigma = \{ Y \mapsto X \}$). This clause
then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ using \infname{ER}.

Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
mathematicians do when they ``replace equals by equals''. Name paramodulation is
also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
states that an equation $s \eq t$ holds under condition $C$ (or more precisely,
its negation). Similarly, the right premise asserts that a (dis)equation holds
under a condition. The superposition rule simply concatenates the conditions for
both premises and replaces equals by equals in two main inference terms. For
example given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and
$\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
between their last literals results in $\neg \cst{even}(X) \llor \neg
\cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.

>>>>

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, first-order logic and higher-order logic. Then,
    we explain the clausal structure which is the backbone of many calculi for
    automated provers. We finish with the description of the superposition
    calculus. As this thesis discusses practical aspects of theorem proving, we
    define only the fundamental notions, while more advanced notions are
    intuitively described with references to rigorous definitions. The text of
    this chapter is partly based on the preliminaries sections of the
    publications listed in Chapter \ref{ch:intro}.
\end{abstract}

@@

Terms are used to build \emph{atoms}. An atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable, then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for every value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic, it does not suffice to assign
values to variables to determine the truth value of the formula. It is also
necessary to interpret the symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

Even though we do not formally define the semantics of logic, we assume the natural
extensions of domain, valuation, interpretation, and model (as defined by Fitting
\cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
of (un)satisfiability and (in)validity are assumed. We write
$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
(i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
$\ourmodel \models N.$  Overloading notation, we write $M \models N$ to denote that
$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$

A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$
denoting the empty tuple. {\em Subterms} and positions are inductively defined
as follows. The term $t$ is a subterm of itself at position $\varepsilon$. If
$s$ is a subterm of $t_i$ at position $p$, then $s$ is a subterm of
$\cst{f}(\tuplen{t})$ at position $i.p$. A {\em context} is a term with zero or
more subterms replaced by a hole $\square$. We write $C[\overline{u}_n]$ for the
term resulting from filling in the holes of a context $C$ with the terms
$\overline{u}_n$, from left to right. We say a term is {\em ground} if it has no
variables. By $u[s]$ we denote that $s$ is a subterm of $u$.

@@

In this thesis we use classic extensional simply typed monomorphic higher-order
logic with the choice operator, but without the description operator, the axiom of infinity,
and the axiom of at least two individuals \cite[Sect.~3.5]{bm-14-automation-ho}.
This logic corresponds to Henkin's extensional type theory. We additionally use
some variations of this logic, clearly stating the differences. Assuming a set of
base types $T$ and a set of symbols $\Sigma$, let us define terms and types.
Types are either base types $\tau \in T$, or function types $\tau \rightarrow
\upsilon$ where both $\tau$ and $\upsilon$ are types. Each symbol $\cst{f} \in
\Sigma$ is assigned a type.

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau \rightarrow
\upsilon$ and $\tau$, respectively, then $s \, t$ is a term of type $\upsilon$.
Lastly, if $x$ is a bound variable of type $\tau$ and $s$ is a term of type
$\upsilon$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau
\rightarrow \upsilon$. The syntactic distinction between free and bound
variables gives rise to \emph{loose bound variables} \cite{tn-93-patterns},
bound variables that appear without enclosing $\lambda$ binder  (e.g., $y$ in
$\lamx{y \, \cst{a}}$). Note that there is no requirement that a
symbol is not applied to terms of Boolean type as in first-order logic.
Furthermore, all logical connectives are symbols present in the set $\Sigma$.
This means that higher-order logic does not distinguish between terms and
formulas. For example, $\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand
\cst{q})$ are well-formed terms of higher-order logic. For convenience, we still
call terms of Boolean type \emph{formulas}. Our logic also supports the Hilbert
choice operator $\varepsilon$. Intuitively, this operator chooses an arbitrary value that
satisfies a predicate and it is axiomatized as follows: $\iforall p. \, (\iexists
x. \, p \, x) \iimplies p \, (\varepsilon \, p)$.

\looseness=-1
With $\alpha$-conversion we assume the rule $(\lambda x. \, s) \longrightarrow
\lambda y. \, \{x \mapsto y\}(s)$ where $y$ does not occur as a loose bound variable in $s$
and $y$ is not bound in $s$. Application of a substitution to a term,
implicitly $\alpha$-renames bound variables to avoid capture.  For example, $\{X
\mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$. The
$\beta$-reduction rule, roughly speaking, corresponds to passing arguments in a
function call: $(\lambda x. s) \, t \longrightarrow \{x \mapsto t\} (s)$, where
the bound variables in $s$ are renamed to avoid capture. Lastly,
$\eta$-reduction is defined as $\lambda x.\, s \, x \longrightarrow s$ under the
condition that $x$ is not loose bound in $x$. We write $s \equi t$ if terms
$s$ and $t$ are equal modulo $\alpha\beta\eta$-conversion. Terms that are
convertible with respect to any of the rules form equivalence classes. Thus, we
also say that $\alpha$($\beta\eta$)-convertible terms are
$\alpha$($\beta\eta$)-equivalent.

@@

We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$
and $\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$,
omitting the length $n \geq 0$ when it is not important or it can be inferred. Every
$\beta$-reduced term can be written as $\lam{\tuple{x}{m}}{a \, \tuplen{t}}$,
where $a$ is not an application; we call $a$ the \emph{head} of the term. By
convention, $a, b$, and $\zeta$ denote heads. If $a$ is a free variable, we call the term
\emph{flex}; otherwise, the term is \emph{rigid}.

Deviating from the standard notion of higher-order subterm, we define subterms
on $\beta$-reduced terms as follows: a term $t$ is a subterm of $t$ at position
$\varepsilon$. If $s$ is a subterm of $u_i$ at position $p$, then $s$ is a
subterm of $a\;\overline{u}_n$ at position $i.p$. If $s$ is a subterm of $t$ at
position $p$, then $s$ is a subterm of $\lambda x.\, t$ at position $1.p$. Our
definition of subterm gracefully generalizes the corresponding first-order
notion: $\cst{a}$ is a subterm of $\cst{f} \, \cst{a} \, \cst{b}$ at position 1,
but $\cst{f}$ and $\cst{f} \, \cst{a}$ are not subterms of $\cst{f} \, \cst{a}
\, \cst{b}$. We say a term is ground if it has no free variables, and closed
if it has no loose bound variables.

Throughout this thesis, we consider completeness of higher-order calculi only
with respect to Henkin semantics \cite{bm-14-automation-ho}. Note that no
complete, consistent calculus can exist for higher-order calculi with standard (full)
semantics.

@@

The standard resolution and superposition calculi do not work directly on formulas, but
on their normal forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed into a set of clauses. For propositional
and first-order logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}. In higher-order logic,
formulas are first-class citizens and higher-order calculi can in general work
with nonnormalized problem, in its original form. However, resolution and
superposition-based higher-order calculi still perform best when the problem is clausified
using transformation similar to the one used in the first-order case. We have analyzed
different approaches to clausification in higher-order context \cite{bbtv-21-full-ho-sup}.

\looseness=-1
An {\em equation} $s \eq t$ corresponds to an unordered pair of terms. A {\em
literal} $l$ is an equation $s \eq t$ or its negation (disequation) $s \noteq
t$. A clause $C$ is a finite multiset of literals, interpreted and written
disjunctively:\ $l_1 \llor \cdots \llor l_n$. Free variables of clauses are
implicitly universally quantified. Note that clause-level operators ($\llor, \eq, \noteq$) are not typeset in bold
to distinguish them from term-building symbols ($\ior, \ieq, \ineq, \iand, \iforall, \ldots$) which are typeset in bold. In standard,
non-clausal first-order logic, an atom is either predicate or equational. For
uniformity, and to stay close to the implementation, we encode predicate atoms
as equations with $\itrue$. Negative literals are encoded as disequations. For
example, $\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\cst{even}(x) \not\eq \itrue$. To lighten the
notation, we sometimes write the predicate literals in nonencoded form.
Applying a substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. We say
$\sigmacl{C}$ is a ground instance of $C$ if $\sigmacl{C}$ has no free variables.
\pagebreak[2]

@@

This calculus works on problems in clausified form and proves the problem by
refuting its negation. In practice this means that, to apply superposition to a
given problem, we first must negate its conjecture and clausify the whole
problem using some of the clausification algorithms
\cite{nw-01-small-cnf,rsv-16-vcnf}. The calculus then applies its inference
rules to the clauses, adding results of the inferences to the working set of
clauses. As this process is quite prolific it also uses an order on terms which gives an indication of
term's ``simplicity'' to determine
if some clauses can be simplified or eliminated from the working set.
Superposition is refutationally complete: This means that, if results of all
inferences are computed in a systematic, fair manner, the empty clause ($\bot$) will
eventually be derived for a provable problem. As first-order logic is
undecidable, no guarantees are given for unprovable problems.
Before introducing the rules of the calculus, we provide definitions of these and other
background concepts.

\subsection{Term Order and Selection}
\label{sec:pre:order}
Superposition calculus is parameterized by a
term order $\succ$ with the following properties:

\noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
    \textit{Irreflexive} & For all terms $s$, $s \not\succ s$ \\
    \textit{Transitive} & For all terms $s, t, u$,  if $s\succ t$ and $t \succ u$ then $s \succ u$ \\ 
    \textit{Subterm property} & For all terms $s$ and contexts $C$, $C[s] \succ s$ \\
    \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
    \textit{Respects contexts} & For all terms $s,t$ and nonempty contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
    \textit{Ground total} & For all distinct
    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$ \\
    \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
\end{tabular}

The term order is lifted to literals and clauses using the {\em multiset extension} of
$\succ$. The order is extended to multisets as follows
\cite[Sect.~2.5]{bg-01-resolution}. For two multisets $S_1$ and $S_2$, we write $S_1 \succ S_2$ if $S_1 \not= S_2$
and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
> S_2(y)$. We use the notation $S(x)$ to denote the number of occurrences of $x$ in
$S$. To use the  multiset extension, positive literals are represented as $ \{ \{s\},
\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$, which makes the negative literals slightly greater than the positive ones.
Clauses are then represented as 
multisets of such literals.

Two orders that are commonly used in superposition theorem proving are the
Knuth--Bendix order (KBO) \cite[Sect.~5.4.4]{bn-98-tr-and-all-that} and the
lexicographic path order (LPO) \cite[Sect.~5.4.2]{bn-98-tr-and-all-that}.
KBO assigns integer weights to symbols and uses precedence between symbols to
break eventual ties. LPO entirely relies on precedence and inspects term
structures more closely to determine the order.

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
We always specify if the constraint should be interpreted as ordered or unordered.
We say a substitution $\sigma$ is \emph{more general} than $\varrho$ if there exists a substitution $\theta$
such that $\varrho = \theta\sigma$.
A \emph{most general unifier} is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. In first-order logic,
the most general unifier is unique up to variable renaming. For example, $\cst{f}(X) \ unif \cst{f}(Y)$ yields the mgu 
$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

@@

A (higher-order) unifier of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma(s) \equi \sigma(t)$, for all $s \unif t
\in E$. A complete set of unifiers of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier,
when it exists, corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist procedures
that enumerate all elements of the CSU for two terms.

These higher-order concepts gracefully generalize the corresponding first-order
ones. For example, if $s$ at $t$ are first-order, then there must exist a singleton
CSU, i.e., the most general unifier.

@@

Before spelling out the inference rules of standard, first-order superposition,
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote the subterm of $t$ at position $p$ and with $t[p \leftarrow
t']$  we denote the term obtained by replacing this subterm with $t'$. Let \selfun{} be a selection function,
$\succ$ a term order, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within the literals in
\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}.  With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using a horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

@@

\medskip
Let us give some intuition behind how these rules are applied in practice. The
\infname{ER} rule is used to replace equals by equals within a clause.
Consider the formula $\iforall x.\,  x \eq \cst{2} \iimplies \cst{even}(x)$. In
clausal form it is $X \noteq \cst{2} \llor \cst{even}(X)$. Applying the
\infname{ER} rule effectively applies the precondition $X \eq \cst{2}$ to the
conclusion $\cst{even}(X)$ to obtain $\cst{even}(\cst{2})$. The \infname{EF}
rule can be used to find instances of the clause that can be further simplified.
Consider the clause $\cst{f}(X) \eq \cst{a} \llor \cst{f}(Y) \eq \cst{a} \llor
\cst{p}(X, Y)$. Applying \infname{EF} to its first two literals yields $\cst{a}
\noteq \cst{a} \llor \cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ (with $\sigma = \{ Y \mapsto X \}$). This clause
then further simplifies to $\cst{f}(X) \eq \cst{a} \llor \cst{p}(X, X)$ using \infname{ER}.

Lastly, the rules \infname{SP} and \infname{SN} are used to simulate what
mathematicians do when they replace equals by equals. The name paramodulation is
also used for rules that replace equals by equals in calculi preceding superposition.  The left premise intuitively
states that an equation $s \eq t$ holds under condition $C$. Similarly, the right premise asserts that a (dis)equation holds
under a condition. The superposition rule simply concatenates the conditions for
both premises and replaces equals by equals in two main inference terms. For
example given clauses $\neg \cst{even}(X) \llor \cst{f}(X) \eq \cst{a}$ and
$\neg \cst{even}(\cst{b}) \llor \cst{g}(\cst{f}(Y)) \eq \cst{b}$, superposition
between their last literals results in $\neg \cst{even}(Y) \llor \neg
\cst{even}(\cst{b}) \llor \cst{g}(\cst{a}) \eq \cst{b}$.
